On the effects of partying on next-day sports performance

### 2017-06-01 writing/draft.tex
Changed how tables are broken out.

\section{Introduction}
It's not unreasonable to think that game performance can be affected 
(in part) by what takes place off the court. 
Anecdotally, it has been reported that professional sports teams party on 
the road, sometimes to the detriment of the game.\cite{hoch}
Within the NBA, J.R. Smith is infamous for citing his improvement in performance due to a
lack of nightlife when moving to Cleveland from New York.\cite{price,ley}
Matt Barnes recently complained that ``there's no nightlife in Utah'' when 
asked about playing Jazz in the second round of NBA playoffs in 2017.\cite{cestone}
A few interviews with former professional NBA athletes revealed that location
matters more than day of the week, i.e. parties can happen any day of the week,
but they are more likely to occur in select cities.

@@

Studies on effects of alcohol on sports performance in a controlled setting
show that alcohol impairs the nervous system, resulting in decrements in both
cognitive function and motor skill.\cite{shirreffs} In a study on 
the effects of alcohol on recovery of male athletes, it was found that
consumption is correlated with reduced production of testosterone
which ultimately leads to inhibited ability to recover and adapt to exercise.\cite{mjbarnes}
Sleep deprivation has also been shown to strongly impair cognitive and motor performance.\cite{pilcher} Both alcohol and sleep deprivation are well correlated with lifestyles associated
with an active night life.

@@

\paragraph{Logic for using spread data}
In order to assert a causal relation between cities with active nightlife 
and subsequent day sports performance, we need to control for salient
features such as the ability of players on either team in the match-up,
whether any players are injured, the momentum of success each team is carrying
in the season, etc.\cite{anderson} To that end, we utilize spread data. The benefit of
using spread data is that it controls for a lot, but the downside is that
we don't know exactly what it controls for.

For example, if book-keepers
are already aware of the effects of night-life on performance, it could
already be factored into the spread; if this were the case, then our variable
of interest would not appear to have significant explanatory power
after conditioning on spread, in spite of there being
a true effect. Alternatively,
if the house book-keeper systematically over-values
teams which are more likely to be encountered after playing a city with active nightlife,
we could see a spurious effect. 
That said, the amount of money that is at stake and the effort put into creating these spreads 
suggests that they are reasonably reliable. 
So, if we do find an effect in our variable of interest,
we can be confident in our conclusion that there is a causal effect and not some 
other systematic bias.

@@

\paragraph{The problem of an unknown travel schedule} In NBA, we have the difficulty of not observing when teams travel. E.g. if a particular team  plays
in Los Angeles on Friday and another game in New York the following day, we don't
know whether the team traveled on Friday evening or Saturday morning. Heuristics for
travel schedules can be made, but it is hard to obtain reliable team-travel data.
In MLB, on the other hand, each match-up is actually a series of games. E.g.
if the Dodgers play the Giants in San Francisco, they will play several games spread
across several days, each in the same city; in these cases, we can infer that the team
is in fact spending the night in said city.

\subsection{Data}
\paragraph{Lines data} We first look toward 
\href{http://www.basketball-reference.com/leagues/}{basketball-reference.com}
for a listing of game-days by season. With this listing, we queried \href{http://www.covers.com/sports/NBA/matchups?selectedDate=2011-1-01}{covers.com},
for all games occurring on a particular day, and for each we observe the
outcome of each game alongside the line set by a betting house.\cite{hrisports} We obtain
data for seasons 2010-11 through 2016-17. We are missing the first couple months of games 
in the 2010-11 season, and our 2011-12 season has some 229 missing games. In each remaining
season, we have at most two games missing of the $82 \times 30 / 2 = 1,230$ played.
We have a total of 7,829 games in our data-set.

@@

\paragraph{Location and Travel Distance Data} We are also interested
to account for the effects of jet-lag. For each team, we queried
a maps search engine to obtain a longitude and latitude listing.\cite{ggmap}
For each pairwise set of coordinates, we calculated distance between the two
points.\cite{sp} To account for jet-lag, we first computed the final direction 
(bearing) when arriving at the 
second point after starting from the first and following the shortest path on a sphere 
(following a great circle), which yields a direction in degrees. 
We then obtain a measure for east-west travel by looking at the trigonometric-$\sin$ of this angle after converting to radians.
Interacting this with travel-distance gets us a measure of jetlag.\cite{lallensack}

@@

\section{Results}
\subsection{Using spread data}
We find that after accounting for player fatigue incurred during the last game, jetlag and travel fatigue, and recovery time, and home-team effect, that both our discrete indicator and continuous measure of party
have a statistically significant effect on likelihood of meeting the spread.

\subsection{Using performance statistics}
We additionally looked at actual performance metrics instead of relying on spread data. 
This allows us to avoid using a black-box feature, and also pin-point
more exactly what elements of the game are in fact effected after visiting a city with active 
nightlife. This motivates looking at number of points allowed, and number of points scored;
we also considered how rebounds, fouls, and even injuries might be explained (in part) by
a hangover effect.

In particular, we see that the average number of rebounds obtained per minute
of game-play does decrease as a function of being exposed to a city with active night-life.
The method was to first calculate for each player and each game the number of rebounds obtained
per minute of game-play. Then, we subtract from each observation the player's average number of rebounds per minute taken across the entire season.
This is an ad-hoc approach that cannot be used for betting, since at game-time we are relying on future data for use in our average statistic which we must calculate and sweep out from each observation.

@@

In general, we can replicate our analysis and use the same resources to obtain our data,
 but there are also some notable differences.
In the MLB, money-lines
are used instead of point-spreads. So as opposed to predicting
the likelihood of meeting the spread, we predict the probability
a team would win against their opponent as both a function of the book-makers'
odds of winning and our party variable. 
Additionally, our methodology changes ever so slightly: when creating our
continuous measure of nightlife, we interact the variable with weekend. The logic
is that because there are so many more games played in MLB, the weekend effect is more salient.
One of the last major differences between NBA and MLB games is that MLB games
are played in series, so the same match-up repeated multiple times. This remedies
the problem of an unknown travel schedule we faced with NBA because we are sure 
teams stay overnight in the same location within a series.

% Table created by stargazer v.5.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu % Date and time: Mon, May 29, 2017 - 11:56:17 PM
\hspace{-105pt}
\begin{tabular}{@{\extracolsep{5pt}}lcccc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]   & \multicolumn{4}{c}{\textit{Dependent variable:}} \\  \cline{2-5}  \\[-1.8ex] & \multicolumn{2}{c}{NBA: Meet Spread} & Demeaned Reb/Min & MLB: Win Game \\  \\[-1.8ex] & \multicolumn{2}{c}{\textit{logistic}} & \textit{OLS} & \textit{logistic} \\  \\[-1.8ex] & (1) & (2) & (3) & (4)\\  \hline \\[-1.8ex]  party.discrete & $-$0.406$^{***}$ (0.133) &  &  &  \\    party.continuous &  & $-$0.053$^{**}$ (0.022) & $-$0.001$^{*}$ (0.0005) & $-$0.035$^{*}$ (0.020) \\    lag.change.position & 0.0001 (0.002) & 0.0001 (0.002) & 0.00003 (0.0001) &  \\    odds &  &  &  & 2.602$^{***}$ (0.162) \\    log(travel.dist+1) & $-$0.017 (0.025) & $-$0.015 (0.025) & 0.001 (0.001) & $-$0.001 (0.005) \\    east-west.travel.dir& $-$0.114 (0.230) & $-$0.101 (0.230) &  &  \\    nhours.last.game & 0.001 (0.001) & 0.001 (0.001) & 0.00003 (0.00002) &  \\    hour(date) & 0.007 (0.012) & 0.005 (0.012) &  &  \\    weekend &  &  &  & 0.040 (0.036) \\    (team == location) & $-$0.011 (0.043) & $-$0.001 (0.043) &  & 0.025 (0.034) \\    log(travel.dist+1):ew & 0.018 (0.035) & 0.016 (0.035) &  &  \\    ndays.last.game &  &  &  & 0.014 (0.019) \\    Constant & $-$0.077 (0.374) & $-$0.041 (0.374) & $-$0.007$^{*}$ (0.004) & $-$1.331$^{***}$ (0.080) \\   \hline \\[-1.8ex]  Observations & 10,000 & 9,979 & 44,190 & 25,215 \\  R$^{2}$ &  &  & 0.0003 &  \\  Adjusted R$^{2}$ &  &  & 0.0002 &  \\  Log Likelihood & $-$6,925.180 & $-$6,912.606 &  & $-$17,274.450 \\  Akaike Inf. Crit. & 13,868.360 & 13,843.210 &  & 34,562.900 \\  Residual Std. Error &  &  & 0.105 (df = 44185) &  \\  F Statistic &  &  & 3.064$^{**}$ (df = 4; 44185) &  \\  \hline  \hline \\[-1.8ex]  \textit{Note:}  & \multicolumn{4}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\  \end{tabular}

\subsubsection{Betting performance}
We also validate our model by seeing how well we can perform against the
market.
\begin{figure}[!h]
  \centering
  \includegraphics{mlb_bets_by_season}
  \caption{Betting performance by season. In each season, The 
    circled point represents the worst out of pocket expense incurred during the season.
    We first train a model on historical data then use said model to predict on ``current year''
    data, ploting profits as a function of time over the ``current'' season. For each relevant
    game, we place a bet if our model instructs us of odds sufficiently different from what
    the betting house suggests. We account for the bookmaker's cut. E.g. in the early 2016 season,
    at its worst we are out of pocket \$750 dollars. By season end, we have recouped all of
    our initial investment plus an additional \$11.5k.}
  \label{bettingperf}
\end{figure}

>>>>

\section{Introduction}
It's not unreasonable to think that game performance can be affected 
(in part) by what takes place off the court. 
Anecdotally, it has been reported that professional sports teams party on 
the road, sometimes to the detriment of the game.
Within the NBA, J.R. Smith is infamous for citing his improvement in performance due to a
lack of nightlife when moving to Cleveland from New York.\cite{price,ley}
Matt Barnes recently complained that ``there's no nightlife in Utah'' when 
asked about playing Jazz in the second round of NBA playoffs in 2017.\cite{cestone}
A few interviews with former professional NBA athletes revealed that location
matters more than day of the week, i.e. parties can happen any day of the week,
but they are more likely to occur in select cities.

@@

Studies on effects of alcohol on sports performance in a controlled setting
show that alcohol impairs the nervous system, resulting in decrements in both
cognitive function and motor skill.\cite{shirreffs} In a study on 
the effects of alcohol on recovery of male athletes, it was found that
consumption is correlated with reduced production of testosterone
which ultimately leads to inhibited ability to recover and adapt to exercise.\cite{mjbarnes}
Sleep deprivation has also been shown to strongly impair cognitive and motor performance.\cite{pilcher} Both alcohol and sleep deprivation are well correlated with nightlife.

@@

\paragraph{Logic for using spread data}
In order to assert a causal relation between cities with active nightlife 
and subsequent day sports performance, we need to control for salient
features such as the ability of players on either team in the match-up,
whether any players are injured, the momentum of success each team is carrying
in the season, etc. To that end, we utilize spread data.\cite{anderson} The benefit of
using spread data is that it controls for a lot, but the downside is that
we don't know exactly what it controls for.

For example, if book-keepers
are already aware of the effects of night-life on performance, it could
already be factored into the spread; if this were the case, then our variable
of interest would not appear to have significant explanatory power
after conditioning on spread, in spite of there being
a true effect. Alternatively,
if the house book-keeper systematically mis-values
teams which are more likely to be encountered after playing a city with active nightlife,
we could see a spurious effect. 
That said, the amount of money that is at stake and the effort put into creating these spreads 
suggests that they are reasonably reliable. 
So, if we do find an effect in our variable of interest,
we can be reasonably confident in our conclusion that there is a causal effect and not some 
other systematic bias.

@@

\paragraph{The problem of an unknown travel schedule} In NBA, we have the difficulty of not observing when teams travel. E.g. if a particular team  plays
in Los Angeles on Friday and another game in New York the following day, we don't
know whether the team traveled on Friday evening or Saturday morning and hence don't know
which city they ``visited'' on Friday night. Heuristics for
travel schedules can be made, but it is hard to obtain reliable team-travel data.
In MLB, on the other hand, each match-up is actually a series of games. E.g.
if the Dodgers play the Giants in San Francisco, they will play several games spread
across several days, each in the same city; in these cases, we can infer that the team
is in fact spending the night in said city.

\subsection{Data}
\paragraph{Lines data} We first look toward 
\href{http://www.basketball-reference.com/leagues/}{basketball-reference.com}
for a listing of game-days by season. With this listing, we queried \href{http://www.covers.com/sports/NBA/matchups?selectedDate=2011-1-01}{covers.com},
for all games occurring on a particular day, and for each we observe the
outcome of each game alongside the line set by a betting house. We interpret the lines
according to a standard guide.\cite{hrisports} We obtain
data for seasons 2010-11 through 2016-17. There are exactly 30 teams in the NBA,
each plays 82 games per season.
We are missing the first couple months of games 
in the 2010-11 season, and our 2011-12 season has 229 missing games. In each remaining
season, we have at most two games missing of the $82 \times 30 / 2 = 1,230$ played.
We have a total of 7,829 games in our data-set.

@@

\paragraph{Location and Travel Distance Data} We are also interested
to account for the effects of jet-lag. For each team, we queried
a maps search engine to obtain a longitude and latitude describing their
stadium location.\cite{ggmap}
For each pairwise set of coordinates, we calculated distance between the two
points.\cite{sp} To account for jet-lag, we first computed the final direction 
(bearing) when arriving at the 
second point after starting from the first and following the shortest path on a sphere 
(following a great circle), which yields a direction in degrees. 
We then obtain a measure for east-west travel by looking at the trigonometric-$\sin$ of this angle after converting to radians.
Interacting this with travel-distance gets us a measure of jetlag.\cite{lallensack}

@@

\section{Results}
\subsection{Using spread data}
We find that after accounting for player fatigue incurred during the last game, jetlag and travel fatigue, recovery time, and home-team effect that both our discrete indicator and continuous measure of party
have a statistically significant effect on likelihood of meeting the spread.
In laymens terms, teams under perform book-maker's expectations after playing in cities 
with active nightlife.
This supports the hypothesis that players enjoy nightlife systematically in certain
cities and that a ``hangover'' effect is felt on the court the next day.

\begin{tabular}{@{\extracolsep{5pt}}lcc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]   & \multicolumn{2}{c}{\textit{Dependent variable:}} \\  \cline{2-3}  \\[-1.8ex] & \multicolumn{2}{c}{Meet the Spread} \\  \\[-1.8ex] & (1) & (2)\\  \hline \\[-1.8ex]    party.discrete & $-$0.433$^{***}$ (0.118) &  \\    party.continuous &  & $-$0.053$^{**}$ (0.023) \\    lag.change.position & $-$0.0001 (0.002) & 0.00005 (0.002) \\    log(travel.dist + 1) & $-$0.019 (0.026) & $-$0.015 (0.026) \\    east-west.travel.dir & $-$0.151 (0.231) & $-$0.116 (0.231) \\    nhours.last.game & 0.001 (0.001) & 0.001 (0.001) \\    hour(date) & 0.006 (0.013) & 0.005 (0.013) \\    (team == location) & $-$0.007 (0.043) & 0.004 (0.043) \\    log(travel.dist + 1):ew & 0.022 (0.036) & 0.018 (0.036) \\    Constant & $-$0.019 (0.378) & $-$0.037 (0.378) \\   \hline \\[-1.8ex]  Observations & 9,844 & 9,823 \\  Log Likelihood & $-$6,814.705 & $-$6,804.350 \\  Akaike Inf. Crit. & 13,647.410 & 13,626.700 \\  \hline  \hline \\[-1.8ex]  \textit{Note:}  & \multicolumn{2}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\  \end{tabular}

We mention that it's not surprising that features describing jet-lag and recovery time
are not significant: such salient information has already been taken into account in
setting the spread.
In the next section, we explore which specific areas of gameplay are affected.

\subsection{Using performance statistics}
We additionally looked at actual performance metrics instead of relying on spread data. 
This allows us to avoid using a black-box feature, and also pin-point
more precisely what elements of the game effected by visiting a nightlife city. This motivates looking at number of points allowed, and number of points scored; we also considered how rebounds, fouls, and even injuries might be 
explained (in part) by a hangover effect.

\paragraph{Rebounds: the ``hustle'' statistic}
In particular, we see that the average number of rebounds obtained per minute
of game-play does decrease as a function of being exposed to a city with active night-life. Rebounds are often considered a ``hustle-statistic'' in the sense 
that anecdotally, players who work harder on the court and expend energy boxing out
other players will have higher rebound rates. If this common knowledge is accepted,
our analysis suggests that players are working less hard following travel to a 
city with active nightlife.

The method was to first calculate for each player and each game the number of rebounds obtained
per minute of game-play. Then, we subtract from each observation the player's average number of rebounds per minute taken across the entire season.
This is an ad-hoc approach that cannot be used for betting, since at game-time we are relying on future data for use in our average statistic which we must calculate and sweep out from each observation.

\paragraph{The hangover correlates with allowing more points}
Similarly, common knowledge also suggests that the overall defense of a team
is effected strongly by individual player effort. It seems likely that a hangover effect
would be more prevalent on the defensive aspects of gameplay. To that end, we examine points allowed and points scored as a function of exposure to nightlife.
It's worth noting that travel to party cities is correlated with allowing
more points on the subsequent day of gameplay, but the null hypothesis
can't be rejected for points scored.

It's also interesting to see that after visiting LA or NY, for any given
team their next day performance is strongly correlated with allowing the other team to
score two and a half points more. 
It's even more interesting to note that points spreads are on a similar order of magnitude,  
i.e. this could often be the difference between meeting the spread or not.

\hspace{-80pt}
\begin{tabular}{@{\extracolsep{5pt}}lccc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]   & \multicolumn{3}{c}{\textit{Dependent variable:}} \\  \cline{2-4}  \\[-1.8ex] & demeaned.pts.admitted & team.pts.admitted & team.pts.scored \\  \\[-1.8ex] & (1) & (2) & (3)\\  \hline \\[-1.8ex]   party.discrete & 2.603$^{***}$ (0.542) &  &  \\    party.continuous &  & 0.486$^{***}$ (0.141) & $-$0.003 (0.138) \\    lag.change.position &  & 0.267$^{***}$ (0.016) & 0.285$^{***}$ (0.016) \\    log(travel.dist+1) &  & $-$0.429$^{**}$ (0.191) & 0.140 (0.187) \\    nhours.last.game &  & 0.012$^{*}$ (0.007) & 0.016$^{**}$ (0.007) \\    Constant & $-$0.072 (0.090) & 70.598$^{***}$ (2.266) & 64.914$^{***}$ (2.218) \\   \hline \\[-1.8ex]  Observations & 15,658 & 6,426 & 6,426 \\  R$^{2}$ & 0.001 & 0.045 & 0.051 \\  Adjusted R$^{2}$ & 0.001 & 0.045 & 0.051 \\  Residual Std. Error & 11.157 (df = 15656) & 11.937 (df = 6421) & 11.688 (df = 6421) \\  F Statistic & 23.077$^{***}$ (df = 1; 15656) & 75.897$^{***}$ (df = 4; 6421) & 86.787$^{***}$ (df = 4; 6421) \\  \hline  \hline \\[-1.8ex]  \textit{Note:}  & \multicolumn{3}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\  \end{tabular}

@@

In general, we can replicate our analysis and use the same resources to obtain our data,
 but there are also some notable differences.
In the MLB, money-lines
are used instead of point-spreads. So as opposed to predicting
the likelihood of meeting the spread, we predict the probability
a team would win against their opponent as both a function of the book-makers'
odds of winning and our party variable. 
Additionally, our methodology changes ever so slightly: when creating our
continuous measure of nightlife, we interact the variable with weekend. The logic
is that because there are so many more games played in MLB, the weekend effect is more salient.
One of the last major differences between NBA and MLB games is that MLB games
are played in series, so the same match-up is repeated multiple times. This remedies
the problem of an unknown travel schedule we faced with NBA because we are sure 
teams stay overnight in the same location within a series.

\begin{tabular}{@{\extracolsep{5pt}}lc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]   & \multicolumn{1}{c}{\textit{Dependent variable:}} \\  \cline{2-2}  \\[-1.8ex] & (team.score $ > $ opponent.score) \\  \hline \\[-1.8ex]   party & $-$0.035$^{*}$ (0.020) \\    odds & 2.602$^{***}$ (0.162) \\    I(team == location) & 0.025 (0.034) \\    ndays.lgame & 0.014 (0.019) \\    log(travel.dist + 1) & $-$0.001 (0.005) \\    weekend & 0.040 (0.036) \\    Constant & $-$1.331$^{***}$ (0.080) \\   \hline \\[-1.8ex]  Observations & 25,215 \\  Log Likelihood & $-$17,274.450 \\  Akaike Inf. Crit. & 34,562.900 \\  \hline  \hline \\[-1.8ex]  \textit{Note:}  & \multicolumn{1}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\  \end{tabular}

\subsubsection{Betting performance}
We also validate our MLB model by seeing how well we can perform against the
market.
\begin{figure}[!h]
  \centering
  \includegraphics{mlb_bets_by_season}
  \caption{Betting performance by season. In each season, The 
    circled point represents the worst out of pocket expense incurred.
    We first train a model on historical data then use said model to predict on ``current year''
    data, plotting profits as a function of time over the ``current'' season. For each relevant
    game, we place a bet of \$100 if the expected value of the bet is positive after accounting f
    or the bookmaker's cut. E.g. in the     early 2016 season,
    at its worst we are out of pocket \$750 dollars. By season end, we have     recouped all of
    our initial investment plus an additional \$11.5k.}
  \label{bettingperf}
\end{figure}
